---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>% glimpse
```

**Observations**:

- Class, Sex, Age, Survived, n

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- There were a total of 2201 people that boarded the Titanic according to this dataset.
- Are there any differences?
  - Yes, according to the Wikipedia article, there were 885 crew members and 1317 Passengers, which add to 2202.
- If yes, what might account for those differences?
  - One person could have purchased a ticket but not have gotten on. If ticket sales were counted for the passenger headcount, it might account for the difference.
  - The dataset might not include cases where it is unknown whether the person survived.

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic[df_titanic$Survived == "Yes", ] %>%
  ggplot(aes(x = Sex, y = n, fill = Sex)) +
  geom_bar(stat = "identity") +
  facet_grid(~ Class)
```

**Observations**:

- For 1st and 2nd Class passengers, significantly more females are rescued. For 3rd class, while there are more females that were rescued, the difference is not as big as 1st and 2nd class. For Crew, the amount of males rescued was much bigger than females, more so than the difference of 1st and 2nd class.



# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop[df_prop$Survived == "Yes", ] %>%
  ggplot(aes(x = Sex, y = Prop, fill = Sex)) +
  geom_bar(stat = "identity") +
  facet_grid(~ Class)
```

**Observations**:


- Across the Class, females had a better chance of getting rescued, even for crew, which had more males that were actually rescued.
- Is there anything *fishy* going on in your plot?
  - The proportions add up to more than 1.

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_q5 <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    total = sum(n),
    prop = n/total
    )

df_q5[df_q5$Survived == "Yes", ] %>%
  ggplot(aes(x = Sex, y = prop, fill = Sex)) +
  geom_bar(stat = "identity") +
  facet_grid(Age ~ Class)
```

**Observations**:

- Females still had better chances of being rescued, according to this visualization. However, when introducing age as a variable, it is clear that children had much better odds of being rescued compared to adults of the same class and sex.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - In q4, the plot was adding adult and child proportions, which made some values bigger than 1. 

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
