---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

- Write all variable names here

country, continent, year, liefExp, pop, gdpPercap

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% pull(year) %>% max()
year_min <- gapminder %>% pull(year) %>% min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

gapminder %>% 
  filter(year == year_min) %>% 
    ggplot(aes(continent, gdpPercap)) +
    geom_boxplot() +
    scale_y_log10()

```

**Observations**:

- Write your observations here

Oceania has the smallest quartile range, while Asia has the biggest quartile range. I suspected that this was due to the number of countries in each continents, but it turns out that Asia and Europe has just about the same number of countries inside the continent, while Africa has more countries in the continent than Asia.

**Difficulties & Approaches**:

- Write your challenges and your approach to solving them

There are multiple countries in each continent, therefore some kind of sum had to be taken. In this case, a boxplot was chosen to visualize, since it not only shows the mean values, but the distribution and outliers.

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>% filter(year == year_min) %>% slice_max(gdpPercap)
gapminder %>% filter(continent == "Europe", year == year_min) %>% slice_max(gdpPercap)
gapminder %>% filter(continent == "Americas", year == year_min) %>% slice_max(gdpPercap)
```

**Observations**:

- Identify the outlier countries from q2
  - (Your response here)
  Kuwait, Switzerland, and the United States stand out as the most unusual in terms of GDP per capita in each continent.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>% filter(year == year_min | year == year_max) %>% 
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United States", "Kuwait", "Switzerland")),
    mapping = aes(color = country),
    size = 2
  ) +
  facet_grid(cols = vars(year)) +
  scale_y_log10()
```

**Observations**:

- Write your observations here

Comparing `year_min` and `year_max`, every continent shows signs of growth in terms of GDP per capita. For the outliers, Switzerland and United states saw growth in GDP per capita between `year_min` and `year_max`, while Kuwait actually saw a decrease. It is also interesting to note that on the log10 scale, the gap that the US had in `year_min` from the upper quartile is still maintained in `year_max`, while the gap for Switzerland and of course, Kuwait, has decreased. 

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gdp_out <- gapminder %>% filter(country == "Kuwait" |
                                   country == "United States" | 
                                   country == "Switzerland")


gdp_out %>% 
  ggplot(aes(x = year, y = gdpPercap, color = country)) +
  geom_line()
```

- (Your notes and observations here) 

The GDP per capita of Kuwait fluctuates wildly before settling into the trend of the other outliers. A question that arises here is that how does the GDP per capita look like for the entire world across the years compared to these three countries?

```{r q5-task2}
## TASK: Your second graph
gdp_percap_mean <- aggregate(gdpPercap ~ year, gapminder, mean)
 
ggplot(gdp_percap_mean, aes(x = year, y = gdpPercap)) +
geom_line() +
geom_line(
    data = gapminder %>% filter(country %in% c("United States", "Kuwait", "Switzerland", "India", "China")),
    mapping = aes(color = country),
  )
```

- (Your notes and observations here)

The worldwide trend is the GDP per capita is increasing as time goes on. I have included China and India here, the two most populated countries in the world. As in the name, GDP per capita, is also dependent on population, it would be interesting to see how the shape of the graphs may change when taking the variable into consideration.

```{r q5-task3}
## TASK: Your third graph
gdp_country <- gapminder %>% filter(country == "Kuwait" |
                                country == "United States" | 
                                country == "Switzerland" |
                                country == "China" |
                                country == "India"
                                )

gdp_country <- gdp_country %>% mutate(gdp = gdpPercap * pop)

gdp_country %>% 
  ggplot(aes(x = year, y = gdp, color = country)) +
  geom_line()
```

- (Your notes and observations here)

When multiplied by population, United States has the biggest GDP by a large margin. It is also interesting to note that while China and India had smaller GDP per capita values, even compared to the mean of the entire world, their GDP itself is larger than that of Kuwait and Switzerland, which both had bigger GDP per capita values than the global mean.
