---
title: "SAT and College Grades"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
---

*Purpose*: How do we apply hypothesis testing to investigating data? In this challenge you'll practice using hypothesis testing tools to make sense of a dataset.

*Reading*:
- [Harvard Study Says SATs Should Be Optional: Here's Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why) (Optional); easy-to-read news article on colleges going SAT-free
- [Norm-Referenced Tests and Race-Blind Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university) (Optional); technical report on relationship between the SAT/ACT and non-academic factors

*Credit*: This is based on a [case study](http://onlinestatbook.com/2/case_studies/sat.html) originally prepared by Emily Zitek, with data collected through the research of Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
library(boot)
```

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

*Background*: Every year about 2 million students take the Scholastic Aptitude Test (SAT). The exam is [controversial](http://www.nea.org/home/73288.htm) but [extremely consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html). There are many claims about the SAT, but we're going to look at just one: Is the SAT predictive of scholastic performance in college? It turns out this is a fairly complicated question to assess---we'll get an introduction to some of the complexities.

# Obtain the Data
<!-- -------------------------------------------------- -->

### __q1__ Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <- read_excel("./data/sat.xls")

## TODO: Do your "first checks"
df_sat %>% glimpse
```

**Observations**:

- Fill in the following "data dictionary"

| Column     | Meaning                  |
|------------|--------------------------|
| `high_GPA` | High school GPA          |
| `math_SAT` | Math SAT score           |
| `verb_SAT` | Verbal SAT score         |
| `comp_GPA` | Computer science GPA     |
| `univ_GPA` | Overall university GPA   |

- What information do we have about these students?
  - (Your response here) Their major, SAT score, GPA from high school, computer science, and university, and that they are all from the same university.
- What kinds of information *do we not have* about these students?
  - (Your response here) Everything else about them, including high school activities.
- Based on these missing variables, what possible effects could be present in the data that we would have *no way of detecting*?
  - (Your response here) Academic dishonesty, and students taking different SAT exams could pose an issue.

# Analysis with Hypothesis Testing
<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the first based on hypothesis testing of correlation coefficients, and the second based on fitting a regression model and interpreting the regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)
```

## View 1: Correlations
<!-- ----------------------------------------------------------------------- -->

### __q2__ Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
## TODO:
df_composite %>% 
  ggplot(aes(x = high_GPA, y = univ_GPA)) + 
  geom_point(aes(color = both_SAT)) +
  scale_colour_gradient(low = "red", high = "green")
```

**Observations**:

- What relationship do `univ_GPA` and `both_SAT` exhibit?
  - (Your response here) As seen from the color groupings, the red dots tend to be on the lower part of the graph, while the greener dots tend to be higher on the graph. This means that those with lower `both_SAT` values tend to have a lower `univ_GPA`.
- What relationship do `univ_GPA` and `high_GPA` exhibit?
  - (Your response here) Points tend to go from the bottom left to top right, which means that the higher the `high_GPA`, the higher the `univ_GPA`.

### Hypothesis Testing with a Correlation Coefficient
<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation coefficient. The
idea is to set our null hypothesis to the case where there is no correlation,
and test to see if the data contradict that perspective. Formally, the null (H0)
and alternative (HA) hypotheses relating to a correlation coefficient between
two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the assumption that `X, Y` are both normally distributed. First, let's check to see if this assumption looks reasonable for our data.

### __q3__ Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
# df_composite %>% 
#   ggplot() +
#   geom_histogram(aes(x = both_SAT)) +
#   geom_histogram(aes(x = high_GPA)) +
#   geom_histogram(aes(x = univ_GPA))

ggplot(df_composite, aes(x = both_SAT)) + geom_histogram()
ggplot(df_composite, aes(x = high_GPA)) + geom_histogram()
ggplot(df_composite, aes(x = univ_GPA)) + geom_histogram()
```

**Observations**:

- To what extent does `both_SAT` look like a normal distribution?
  - (Your response here) The counts taper off on either extremes, however the counts in between are quite evenly spread.
- To what extent does `high_GPA` look like a normal distribution?
  - (Your response here) It looks like there are two humps - as if there were two normal distributions next to each other. Counts taper out on either extremes but at the middle as well. (This is interesting as I heard many of my math teachers talk about how student grades exhibit this exact behavior.)
- To what extent does `univ_GPA` look like a normal distribution?
  - (Your response here) It is similar to the shape of `high_GPA`. However, the hump on the left is significantly smaller than the one on the right.

Keep in mind your findings as you complete q4.

### __q4__ Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA
print("high_GPA vs. univ_GPA:")
cor.test(df_composite$high_GPA, df_composite$univ_GPA)

print(("both_SAT vs. univ_GPA:"))
cor.test(df_composite$both_SAT, df_composite$univ_GPA)
```

**Observations**:

- Which correlations are significantly nonzero?
  - (Your response here) Both `[high_GPA, univ_GPA]` and `[both_SAT, univ_GPA]` have significant nonzero correlation.
- Which of `high_GPA` and `both_SAT` seems to be more strongly correlated with `univ_GPA`?
  - (Your response here) `high_GPA` has a bigger correlation coefficient.
- How do the results here compare with the visual you created in q2?
  - (Your response here) The visual exhibited signs of correlation between both parameters, which agrees with the correlation coefficients.
- Based on these results, what can we say about the predictive capabilities of both `high_GPA` and `both_SAT` to predict `univ_GPA`?
  - (Your response here) One could reasonably predict `univ_GPA` given `high_GPA` or `both_SAT`, or both.

Finally, let's use the bootstrap to perform the same test using *different* assumptions.

### __q5__ Use the bootstrap to approximate a confidence interval for `corr[high_GPA, univ_GPA`. Compare your results---both the estimate and confidence interval---to your results from q4.

```{r q5-task}
## TODO: Use the bootstrap to compute a confidence interval for corr[high_GPA, univ_GPA]
high_univ <- df_composite[, c("high_GPA", "univ_GPA")]

corr_func <- function(data) {
  return(cor.test(data[, 1], data[, 2]))
}

high_univ_boot <- bootstraps(data = high_univ, times = 1000) %>% 
  mutate(df = map(splits, ~ analysis(.x)))

corr_boot <- high_univ_boot$df

df_boot <- as.data.frame((do.call(rbind, corr_boot)))

cor.test(df_boot$high_GPA, df_boot$univ_GPA)
```

**Observations**:

- How does your estimate from q5 compare with your estimate from q4?
  - (Your response here) The correlation value has not changed significantly from q4.
- How does your CI from q5 compare with your CI from q4?
  - (Your response here) The bootstrapped CI is much smaller than the one from q4.

*Aside*: When you use two different approximations to compute the same quantity and get similar results, that's an *encouraging sign*. Such an outcome lends a bit more credibility to the results.

## View 2: Modeling
<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study the relationship among more variables we can instead use a fitted model. Using a model, we can also help assess whether it is *worthwhile* to measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model
<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a model, we can express our hypotheses in terms of the model parameters. For instance, if we were interested in whether $X$ has an affect on $Y$, we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on $Y$. Let's apply this idea to relating the variables `univ_GPA` and `high_GPA`. Luckily R has built-in tools to construct a confidence interval on the $\beta$'s in a regression [1]; we'll simply use those tools rather than do it by hand.

### __q6__ Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <- lm(univ_GPA ~ both_SAT, data = df_composite)

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

df_composite %>% 
  ggplot(aes(x = both_SAT, y = univ_GPA)) +
  geom_point() +
  geom_smooth(method = "lm")
```

**Observations**:

- What is the confidence interval on the coefficient of `both_SAT`? Is this coefficient significantly different from zero?
  - (Your response here) The difference in `conf.low` and `conf.high` for `both_SAT` in the model is ~0.0013, which I deem the difference from zero as insignificant.
- By itself, how well does `both_SAT` predict `univ_GPA`?
  - (Your response here) With the plotted model, the trend is definitely present in the regression line. However, there seem to be many points outside the line or even the confidence interval.

Remember from `e-model03-interp-warnings` that there are challenges with interpreting regression coefficients! Let's investigate that idea further.

### __q7__ Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
df_composite %>% 
  ggplot(aes(x = high_GPA, y = univ_GPA)) +
  geom_point() +
  geom_smooth(method = "lm")

df_composite <- df_composite %>% 
  mutate(high_prod = (high_GPA^5)*both_SAT)

df_composite %>% 
  ggplot(aes(x = high_prod, y = univ_GPA)) +
  geom_point() +
  geom_smooth(method = "lm")

fit_both <- lm(univ_GPA ~ high_prod, data = df_composite)

fit_both %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

```

**Observations**:

- How well do these models perform, compared to the one you built in q6?
  - (Your response here)
- What is the confidence interval on the coefficient of `both_SAT` when including `high_GPA` as a predictor?? Is this coefficient significantly different from zero?
  - (Your response here)
- How do the hypothesis test results compare with the results in q6?
  - (Your response here)

## Synthesize
<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### __q8__ Using the results from all previous q's, answer the following questions.

**Observations**:

- Between `both_SAT` and `high_GPA`, which single variable would you choose to predict `univ_GPA`? Why?
  - (Your response here)
- Is `both_SAT` an effective predictor of `univ_GPA`? What specific pieces of evidence do you have in favor of `both_SAT` being effective? What specific pieces of evidence do you have against?
  - (Your response here)

# End Notes
<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more information see this [Wiki article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
